---
title: "lecture 3"
author: "Katrina Bazemore"
date: "2024-09-05"
output: html_document
---

This code was written based on Stata code provided by Dr. Justine Shults and using ChatGPT. The outputs approximate outputs of the Stata code.

Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Change this line to your working directory - where the bmiz datafile is saved
knitr::opts_knit$set(root.dir = "C:/Users/kbazemo/Box/EPID 6210/Data")
```

Loading necessary libraries
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(geepack)
library(qlspack) #to install qlspack, see Install_qlspack.Rmd
library(mice)

#if libraries not already installed, use <install.packages("dplyr")> replacing dplyr with the package name
```

Specifying file containing functions we will use
```{r}
source("../RCode/EPID6210-inR/functions.R")
```

List of functions available in qlspack
```{r}
lsf.str("package:qlspack")
```
Read in the data
```{r}
bmi_data <- read.csv("bmiz_after_renal_transplant.csv")
```

Explore missingness patterns
```{r}
#Get summary stats on number of observations for each subject
summary(bmi_data %>% count(id))[,2]
  #min number of observations per subject is 3, max is 12, average is 6.31
```

```{r}
#Reshape to wide
bmi_data_W <- reshape2::dcast(bmi_data, id ~ month, value.var = "bmiz") 

#Use the mice package to plot missingness patterns
md.pattern(bmi_data_W, plot = T)
```

Fit QLS with AR(1) Structure: using sandwich errors as qlspack only provides 4 types of errors (sandwich robust, approximate jacknife, 1-step jacknife, and fully iterated jacknife)
```{r}
qls_ar1 <- qls(bmiz ~ month + I(month^2), data = bmi_data, id = bmi_data$id, family = "gaussian", correlation = "ar1", std.err = "san.se")
  #for the id argument, have to specify dataset$id, not just id as is typical for other regressions in R
  #for this model I specified sandwich standard errors because there is no model-based option
summary(qls_ar1)

#The function for the estimated correlation matrix still works on qls models - we need to specify the fitting.package as "geepack" because qlspack is built to use functions from geepack
createCorrelationMatrixR(qls_ar1, model.corstr = "ar1", fitting.package = "geepack")
```

Drop some data so that some subjects have only 1 measurement
```{r}
#Remove obs with month>0 and id>40
bmi_edit <- bmi_data %>% arrange(id, month) %>% filter(!(month>0 & id>40))

summary(bmi_edit %>% count(id))[,2]
#Now the min number of observations per person is 1
```

Fit QLS GEE again with AR(1) structure (using sandwich SE as no model-based available)
```{r}
qls_ar1_drop <- qls(bmiz ~ month + I(month^2), data = bmi_edit, id = bmi_edit$id, family = "gaussian", correlation = "ar1", std.err = "san.se")

summary(qls_ar1_drop)
#We can see the the number of groups is still 100, and the max obs per group is still 12

nrow(qls_ar1_drop$data) #number of observations used in modeling

summary(qls_ar1_drop$data %>% count(id))[,2] #Can get summary stats on obs per id directly from the data used in modeling
```

Fit GEE with AR(1) structure on edited data (using sandwich SE as no model-based available)
```{r}
gee_ar1 <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "ar1", data = bmi_edit)
summary(gee_ar1)
#number of clusters is still 100, looks like this does not drop subjects with only one observation as in stata
```

Compare model output
```{r}
#QLS on edited data with sandwich SE
print(summary(qls_ar1_drop)$coefficients)
qls_ar1_drop$geese$alpha
#GEE (geepack) on edited data with sandwich SE
print(summary(gee_ar1)$coefficients)
gee_ar1$geese$alpha

#Results are pretty similar, alphas vary a bit
```

Fit GEE with AR(1) structure on edited data using the GEE package (to see if it performs differently from the geepack package)
```{r}
#gee_arm <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "AR-M", data = bmi_edit)
gee_error <- ("Beginning Cgee S-function, @(#) geeformula.q 4.13 98/01/27
running glm to get initial regression estimate
(Intercept)       month  I(month^2) 
   0.551498    0.026798   -0.000249 
Error in gee::gee(bmiz ~ month + I(month^2), id = id, corstr = AR-M,  : 
  cgee: M-dependence, M=1, but clustsize=1
fatal error for this model")
print(gee_error)
#the GEE package can't handle subjects with only one observation
```

Removing subjects with one observations in the data manually to check the results using the GEE package
```{r}
#Get a list of IDs with only one observation in our edited data
bmi_edit_counts <- bmi_edit %>% count(id) %>% filter(n==1)

#Creating a new dataframe excluding IDs with only only observation in bmi_edit
bmi_edit2 <- bmi_edit %>% filter(!id %in% bmi_edit_counts$id)

#How many observations are we left with
nrow(bmi_edit2)

#Try fitting GEE using GEE package again after removing those with only one observation
gee_arm <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "AR-M", data = bmi_edit2)

P_table(gee_arm, 5) #from the functions.R file
```

Bonus, equivalent to xtset in R
```{r}
library(plm)
plm::is.pbalanced(bmi_data$id, bmi_data$month)
```







