---
title: "lecture4"
author: "Katrina Bazemore"
date: "2024-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Change this line to your working directory - where the bmiz datafile is saved
knitr::opts_knit$set(root.dir = "C:/Users/kbazemo/Box/EPID 6210/Data")
```

Loading necessary libraries
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(geepack)
library(qlspack)

#if libraries not already installed, use <install.packages("dplyr")> replacing dplyr with the package name
```

Specifying file containing functions we will use
```{r}
source("../RCode/EPID6210-inR/functions.R")
```


Read in the transplant data and view the first few observations
```{r}
bmi_data <- read.csv("bmiz_after_renal_transplant.csv")
```

Defining missingness patterns and view
```{r}
view_patterns(bmi_data, id, month, 12) #function I wrote for this purpose, from functions.R
```

Fit the most plausible structure (Markov)
```{r}
#Filtering the data to only include observations through month 12
bmi_filter <- bmi_data %>% filter(month <=12) %>% arrange(id, month)

#Running QLS with markov structure
qls_mark <- qls(bmiz ~ month + I(month^2), data = bmi_filter, id = bmi_filter$id, family = "gaussian", correlation = "markov", std.err = "san.se")

#Model summary
summary(qls_mark)
```

What are the two most common patterns of timings of measurements?
```{r}
view_patterns(bmi_data, id, month, 12) 
#the two most common patterns are 0,1,3,6,12 and 0,1,3,12
```

Print the correlation matrices for the two most common patterns of observations
```{r}
#Updated the createCorrelationMatrixR function so it can get estimated correlation from the markov structure (in functions.R)
full_matrix <- createCorrMatrixR(qls_mark, model.corstr = "markov", fitting.package = "qlspack", time_var = "month", use_time_labels = TRUE)

#Wrote function to subset output of createCorrMatrixR (in functions.R)
  #For most common pattern
  subsetCorrMatrix(full_matrix, c("0", "1", "3", "6", "12"))
  
  #For second most common pattern
  subsetCorrMatrix(full_matrix, c("0", "1", "3", "12"))
```


Estimate the correlation between bmiz at month 0 and month 11
```{r}
alpha <- qls_mark$geese$alpha
timegap <- 11-0
estCorr <- alpha^timegap
print(estCorr)
```









