---
title: "lecture5"
author: "Katrina Bazemore"
date: "2024-09-22"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Change this line to your working directory - where the bmiz datafile is saved
knitr::opts_knit$set(root.dir = "C:/Users/kbazemo/Box/EPID 6210/Data")
```

Loading necessary libraries
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(geepack)
library(qlspack)
library(plm)
#if libraries not already installed, use <install.packages("dplyr")> replacing dplyr with the package name
```

Specifying file containing functions we will use
```{r}
source("../RCode/EPID6210-inR/functions.R")
```

Read in the data
```{r}
bmi_data <- read.csv("bmiz_after_renal_transplant.csv")
```

Filter data
```{r}
#Filter out observations after month 1, and then those with only one observation
bmi_data_edit <- bmi_data %>%
  filter(month == 0 | month == 1) %>% 
  arrange(id, month) %>%
  group_by(id) %>%
  filter(n() >1) %>%
  ungroup()
```

Is the data balanced?
```{r}
#Check if the data is balanced given the id and time variables
plm::is.pbalanced(bmi_data_edit$id, bmi_data_edit$month)
```

Describe the data
```{r}
#How many participants are present?
length(unique(bmi_data_edit$id))

#What are the patterns of observation, is there missingness?
view_patterns(bmi_data_edit, id, month, 1)
```
Convert month to a factor so it plays nice with geepack
```{r}
bmi_data_edit$month <- factor(bmi_data_edit$month, levels=c("0", "1"), labels = c("baseline", "month1"))
```


geepack has a QIC function. It does not work on QLS objects
```{r}
gee_ar1 <- QIC(geeglm(bmiz ~ month, id = id, corstr = "ar1", data = bmi_data_edit))
gee_exch <- QIC(geeglm(bmiz ~ month, id = id, corstr = "exchangeable", data = bmi_data_edit))
#gee_unstruc <- QIC(geeglm(bmiz ~ month, id = id, corstr = "unstructured", data = bmi_data_edit))
  #unstructured won't run due to contrast errors

rbind(gee_ar1, gee_exch)
```

Filter data to months 0, 1, 48
```{r}
#Filter out observations after month 1, and then those with only one observation
bmi_data_edit2 <- bmi_data %>%
  filter(month == 0 | month == 1 | month == 48) %>% 
  arrange(id, month) 

#Check if the data is balanced given the id and time variables
plm::is.pbalanced(bmi_data_edit2$id, bmi_data_edit2$month)

#How many participants are present?
length(unique(bmi_data_edit2$id))

#What are the patterns of observation, is there missingness?
view_patterns(bmi_data_edit2, id, month, 48)
```

Examine model fit when keeping months 0, 1, 48
```{r}
gee_ar1_2 <- QIC(geeglm(bmiz ~ month, id = id, corstr = "ar1", data = bmi_data_edit2))
gee_exch_2 <- QIC(geeglm(bmiz ~ month, id = id, corstr = "exchangeable", data = bmi_data_edit2))
gee_unstruc_2 <- QIC(geeglm(bmiz ~ month, id = id, corstr = "unstructured", data = bmi_data_edit2))

rbind(gee_ar1_2, gee_exch_2, gee_unstruc_2)
```
Results for QLS Markov, GEE unstructured, GEE exchangeable
```{r}
gee_ar1_2 <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "ar1", data = bmi_data_edit2)
gee_exch_2 <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "exchangeable", data = bmi_data_edit2)
gee_unstr_2 <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "unstructured", data = bmi_data_edit2)
qls_mark <- qls(bmiz ~ month + I(month^2), data = bmi_data_edit2, id = bmi_data_edit2$id, family = "gaussian", correlation = "markov", std.err = "san.se")

summary(gee_ar1_2)
summary(gee_exch_2)
summary(gee_unstr_2)
summary(qls_mark)
```




