---
title: "Lecture2_2024_Overview: In R"
author: "Katrina Bazemore, MPH"
date: "2024-08-29"
output: html_document
---

This code was written based on Stata code provided by Dr. Justine Shults, using reference code written by Dr. Erin Schnellinger, and using ChatGPT. The outputs approximate outputs of the Stata code.

Set up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Change this line to your working directory - where the bmiz datafile is saved
knitr::opts_knit$set(root.dir = "C:/Users/kbazemo/Box/EPID 6210/Data")
```

Loading necessary libraries
```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(geepack)
library(gtsummary)

#if libraries not already installed, use <install.packages("dplyr")> replacing dplyr with the package name
```

Specifying file containing functions we will use
```{r}
source("../RCode/EPID6210-inR/functions.R")
#source("functions.R")
```


Read in the transplant data and view the first few observations
```{r}
bmi_data <- read.csv("../../Data/bmiz_after_renal_transplant.csv")

head(bmi_data)
```

Summarize the variables
```{r}
bmi_data %>%
  tbl_summary(
    type = list(everything() ~ "continuous2"),
    statistic = list(all_continuous() ~ c("{mean} ({sd})",
                                          "{min}, {max}")),
    digits = all_continuous() ~ 2,) %>%
  add_n()
```

There are 631 measurements. How many subjects are there?
```{r}
length(unique(bmi_data$id))
```

Descriptive data from lecture 1
```{r}
bmi_data %>% 
  dplyr::group_by(month) %>% 
  dplyr::summarise(Count = n(),
                   Mean=mean(bmiz, na.rm = T))
```

Display the first 10 rows of data
```{r}
bmi_data %>% arrange(id, month) %>% slice(1:10)
```

Regress bmiz on month and month-squared
```{r}
# Regress bmiz on month and month^2
model_bmiz <- lm(bmiz ~ month + I(month^2), data = bmi_data)
summary(model_bmiz)

#Alternatively: summary(lm(bmiz ~ month + I(month^2), data = bmi_data))
#Can also obtain just the coefficients: summary(model_bmiz)$coefficients
```

Obtain and plot the fitted values form the regression model
(R does not have a built in margins command, but the ChatGPT-generated code below seems to work. 
There is also a margins package (https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html) that may replicate some of the functionality of the margins command)
```{r}
margins <- predict(model_bmiz, newdata = data.frame(month = seq(min(bmi_data$month), max(bmi_data$month), by = 1)), interval = "confidence")

# Plotting margins
ggplot(data = data.frame(month = seq(min(bmi_data$month), max(bmi_data$month), by = 1), margins), aes(x = month, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  theme_minimal()

#Save plot as PDF
#ggsave("Lecture2_fittedvalues.png")
#This will save to your working directory
```
Evaluate time lags between consecutive measurements on subjects
```{r}
# Evaluate time lags between consecutive measurements on subjects
bmi_lag_data <- bmi_data %>%
  arrange(id, month) %>%
  group_by(id) %>%
  mutate(lag = month - lag(month, order_by = month))

#Estimate total lags (so we can calculate percents in the next step)
total_lags <- sum(table(bmi_lag_data$lag))

#Tabulate lag values
bmi_lag_data %>%
  dplyr::group_by(lag) %>% 
  dplyr::summarise(Freq = n(),
                   Percent = ((n()/total_lags)*100)
                   )

#Haven't yet figured out how to get the cumulative percent in the table
```

Information about the geeglm command
```{r, eval=FALSE}
help(geeglm)
```

Fit AR1 structure - using the gee package
```{r}
gee_arm <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "AR-M", data = bmi_data)

#Full model summary
#summary(gee_arm)

#Model coefficients
coeffs <- summary(gee_arm)$coefficients
    #Naive S.E is model-based
    #Robust S.E is sandwich-based

#To get a table with coefficients and p-values use the function P_table (from functions.R)
  #Specify the model and number of decimal places to round to
P_table(gee_arm, 4)

#Display the estimated correlation matrix - two options for the gee package
  #Built in matrix
  gee_arm$working.correlation
  
  #Matrix derived using function (from functions.R) - will have to use this function for models estimated with geepack
  createCorrelationMatrixR(gee_arm, model.corstr = "ar1", fitting.package = "gee")
```

Fit AR1 structure - using the geepack package
  Note: the default standard error estimate for geeglm (from geepack) is the robust estimate
  (https://cran.r-project.org/web/packages/geepack/geepack.pdf, page 7)
```{r}
# GEE with AR(1) structure & robust SE
gee_ar1 <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "ar1", data = bmi_data)
summary(gee_ar1)

#Get more information on the model
gee_ar1$geese

#Print just the alpha
gee_ar1$geese$alpha

#Display the estimated correlation matrix - not build in for geepack, have to use function from (functions.R)
createCorrelationMatrixR(gee_ar1, model.corstr = "ar1", fitting.package = "geepack")
```

Fit exchangeable structure - using the gee package
```{r}
gee_exch <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "exchangeable", data = bmi_data)

summary(gee_exch)

P_table(gee_exch, 4)
```

Fit exchangeable structure - using the geepack package
  Again, this defaults to robust SE estimation
```{r}
#GEE with exchangeable structure 
gee_exc <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "exchangeable", data = bmi_data)
summary(gee_exc)

createCorrelationMatrixR(gee_exc, model.corstr = "exchangeable", fitting.package = "geepack")
#Got slightly different estimated correlation structures between the two packages
```

GEE with tri-diagonal structure: neither package has an option for a tri-diagonal correlation structure


Fit unstructured structure - using the GEE package
```{r}
gee_unstruc <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "unstructured", data = bmi_data)

summary(gee_unstruc)

P_table(gee_unstruc, 4)

gee_unstruc$error
#No errors reported

gee_unstruc$iterations
#I think this model was able to converge after 5 iterations
```


Fit unstructured structure - using the geepack package (default = robust SE)
```{r}
gee_unstr <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "unstructured", data = bmi_data)
summary(gee_unstr)

#I think the long list of estimated parameters is telling us there was a failure to converge 

gee_unstr$geese$alpha

#Even though the model did not converge, this function will still output a correlation matrix, but we have many values outside 0-1
createCorrelationMatrixR(gee_unstr, model.corstr = "unstructured", fitting.package = "geepack")
```

Fitting identity structure - using the GEE package
```{r}
gee_ident <- gee::gee(bmiz ~ month + I(month^2), id = id, corstr = "independence", data = bmi_data)

summary(gee_ident)

P_table(gee_ident, 4)
```


Fitting identity structure - using the geepack package (default = robust SE)
```{r}
gee_id <- geeglm(bmiz ~ month + I(month^2), id = id, corstr = "independence", data = bmi_data)
summary(gee_id)

createCorrelationMatrixR(gee_id, model.corstr = "independent", fitting.package = "geepack")
```


